---
title: "Getting started with dist.structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dist.structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dist.structure)
library(algebraic.dist)
```

## What `dist.structure` is

`dist.structure` is a small, principled extension of `algebraic.dist`.
It lets you build and query **random variables with internal
structure**: coherent reliability systems decomposed into components
via a structure function.

Every object produced by `dist.structure` is also a `dist` (from
`algebraic.dist`), so the full distribution algebra (`mean`, `vcov`,
`sampler`, `surv`, `cdf`, `density`) works on it. On top of that,
`dist.structure` exposes structural queries you cannot ask of a plain
distribution: minimal path sets, system signature, Birnbaum
importance, duality, and more.

This vignette is a five-minute tour. Other vignettes go deeper into
specific topics.

## Your first structured distribution

Start with a series system of three independent exponentials:

```{r}
sys <- series_dist(list(
  exponential(0.5),
  exponential(0.3),
  exponential(0.2)
))
sys
```

The object is a dist. Query it like any distribution:

```{r}
algebraic.dist::surv(sys)(1)           # P(system survives past t = 1)
algebraic.dist::cdf(sys)(1)            # P(system fails by t = 1)
set.seed(1)
algebraic.dist::sampler(sys)(5)        # five system-lifetime samples
```

The series-of-exponentials identity `sum(rates) = total rate` holds:
the system is `Exp(sum(rates))`. Confirm:

```{r}
algebraic.dist::surv(sys)(1)
exp(-(0.5 + 0.3 + 0.2) * 1)
```

## Structural queries

Now ask topology questions the base distribution algebra cannot
answer:

```{r}
ncomponents(sys)                       # 3 components
phi(sys, c(1, 1, 0))                   # system functions? No (series needs all)
phi(sys, c(1, 1, 1))                   # Yes
min_paths(sys)                         # the single path: all components
min_cuts(sys)                          # three singleton cuts
system_signature(sys)                  # (1, 0, 0): fails at first failure
```

The structure function `phi` takes a component state vector (0 =
failed, 1 = functioning) and returns 1 if the system is functioning.
For a series system, `phi = AND`: all components must function.

## A richer topology: k-out-of-n

Series and parallel are special cases of k-out-of-n (functions if at
least k of m components function). dist.structure has a shortcut:

```{r}
# 2-out-of-3 of heterogeneous exponentials; closed-form specialization
kofn <- exp_kofn(k = 2, rates = c(1, 2, 3))
ncomponents(kofn)
algebraic.dist::surv(kofn)(1)
system_signature(kofn)                 # (0, 1, 0) for 2-of-3
```

The k-of-n system in dist.structure uses the :G convention: `k`
functioning components are required for the system to function.
Consequently `kofn_dist(k = 1, ...)` is parallel and `kofn_dist(k = m,
...)` is series.

## Sampling, means, and composition

Every distribution generic composes naturally:

```{r}
set.seed(42)
samples <- algebraic.dist::sampler(kofn)(10000)
mean(samples)
```

You can also substitute or compose components:

```{r}
# Swap component 2 with a Weibull.
kofn2 <- substitute_component(kofn, j = 2,
  new_component = weibull_dist(shape = 2, scale = 1))
algebraic.dist::surv(kofn2)(1)
```

## Importance measures

How much does each component matter? dist.structure provides four
complementary importance measures:

```{r}
# Structural importance: fraction of pivotal states for component 1.
structural_importance(kofn, j = 1)

# Birnbaum (reliability) importance: dR/dp_j at given component
# reliabilities.
birnbaum_importance(kofn, j = 1, p = 0.9)

# Criticality importance at time t = 0.5.
criticality_importance(kofn, j = 1, t = 0.5)

# Vesely-Fussell importance at time t = 0.5 (via minimal cut sets).
vesely_fussell_importance(kofn, j = 1, t = 0.5)
```

All four are different answers to "how important is component j?" and
they agree for simple systems but diverge for complex ones. The
[importance measures vignette](importance-measures.html) works through
the distinctions.

## What to read next

- **[Coherent systems](coherent-systems.html)**: the structural
  concepts (phi, min_paths, min_cuts, signature, reliability
  polynomial) in depth, with series, parallel, k-of-n, bridge, and
  consecutive-k examples.
- **[Distribution interface](distribution-interface.html)**: how
  dist_structure inherits from dist, which generics default via
  composition, and when to reach for a specialization (`exp_series`,
  `wei_kofn`) vs the general `coherent_dist`.
- **[Importance measures](importance-measures.html)**: the four
  importance measures applied to a bridge network, including
  time-dependent versions.
- **[Composition](composition.html)**: hierarchical construction via
  `substitute_component` and `compose_systems`.
- **[Non-coherent systems](non-coherent.html)**: cold standby, where
  system lifetime is a sum (not an order statistic), and why this is
  explicitly NOT a dist_structure.

## Ecosystem

`dist.structure` is the shared topology and DGP layer for the
reliability packages in the queelius ecosystem:

- `serieshaz` implements the protocol on `dfr_dist_series` (arbitrary
  dynamic failure rate components).
- `kofn` provides MLE for k-out-of-n systems, delegating its DGP to
  dist.structure.
- `maskedcauses` wraps its exponential-series legacy API over
  `dist.structure::exp_series`.

Any object from these packages that declares `dist_structure` in its
class vector inherits the full protocol automatically: topology
queries, importance measures, composition operators, and the full
`dist` interface.
