---
title: "Implementing a dist_structure subclass"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Implementing a dist_structure subclass}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dist.structure)
```

`dist.structure` is a protocol package: the closed-form constructors
(`exp_series`, `exp_kofn`, ...) and topology shortcuts (`series_dist`,
`kofn_dist`, `bridge_dist`, ...) are reference implementations of an S3
contract that anyone can extend. Downstream packages do exactly that:
`serieshaz` adds a `dfr_dist_series` class for arbitrary dynamic-failure-rate
series systems; `kofn` builds inference machinery on the closed-form
k-out-of-n classes; future packages will add their own.

This vignette walks through the protocol from the implementor's side: what
to provide, what you get for free, when to override defaults for
performance, and how to validate your subclass.

## The contract: four steps

Every `dist_structure` subclass provides:

1. **A class chain** that includes `"dist_structure"`, `"univariate_dist"`,
   and `"dist"`. Almost always also a sub-chain like `"coherent_dist"` if
   the system is coherent.
2. **`ncomponents(x)`**: number of components `m`.
3. **`component(x, j, ...)`**: the j-th component returned as an
   `algebraic.dist::dist` object with parameters baked in. The returned
   dist must be independently evaluable: `surv(component(x, j))(t)`,
   `sampler(component(x, j))(n)`, etc., must all work.
4. **At least one of**:
   - `phi(x, state)`: the structure function, returning 0 or 1 for any
     binary state vector of length `m`.
   - `min_paths(x)`: a list of integer vectors enumerating minimal
     functioning component subsets.

Everything else has a default that derives from those four primitives.
Defaults exist for `min_cuts`, `system_signature`, `critical_states`,
`reliability`, `is_coherent`, `structural_importance`, `system_lifetime`,
`system_censoring`, `dual`, `surv`, `cdf`, and `sampler`.

## A worked example: an alarm system

Consider a fire alarm with `m` redundant smoke sensors and one master
controller. The system **functions** if (i) at least `k` of the `m`
sensors function AND (ii) the master controller functions. This is not
exactly a k-out-of-n system (the master is special) and not exactly a
series-of-(parallel-of-sensors-with-master) (the master gates everything).
We can capture it directly as a custom `dist_structure`.

### Step 1: pick a representation

Components are indexed `1..m` for the sensors and `m+1` for the master.

### Step 2: write the constructor

```{r}
alarm_dist <- function(k, sensor_components, master_component) {
  m_sensors <- length(sensor_components)
  stopifnot(k >= 1L, k <= m_sensors,
            inherits(master_component, "dist"),
            all(vapply(sensor_components,
                       function(d) inherits(d, "dist"),
                       logical(1L))))
  structure(
    list(
      k = as.integer(k),
      m_sensors = as.integer(m_sensors),
      m = as.integer(m_sensors + 1L),
      components = c(sensor_components, list(master_component))
    ),
    class = c("alarm_dist", "dist_structure",
              "univariate_dist", "continuous_dist", "dist")
  )
}
```

The class chain declares membership in `dist_structure` and the
`algebraic.dist` ancestors. We bundle the master into the components
list so the `j`-th component is uniformly accessible.

### Step 3: implement the three required generics

```{r}
ncomponents.alarm_dist <- function(x) x$m

component.alarm_dist <- function(x, j, ...) {
  stopifnot(j >= 1L, j <= x$m)
  x$components[[j]]
}

phi.alarm_dist <- function(x, state) {
  stopifnot(length(state) == x$m)
  sensor_state <- state[seq_len(x$m_sensors)]
  master_state <- state[x$m]
  as.integer(sum(sensor_state) >= x$k && master_state == 1L)
}
```

That's all that's strictly required by the protocol.

### Step 4: validate

```{r}
sensors <- replicate(4, algebraic.dist::exponential(0.5), simplify = FALSE)
master <- algebraic.dist::exponential(0.1)  # master is most reliable
alarm <- alarm_dist(k = 2L, sensors, master)

validate_dist_structure(alarm)
```

If we had forgotten `phi.alarm_dist`, `validate_dist_structure()` would
have stopped with a clear error pointing at the missing primitive
instead of letting the failure surface much later in some default
method's dispatch.

## What you get for free

Every default below is now available on `alarm`:

```{r}
# Number of components
ncomponents(alarm)

# Structure function evaluation
phi(alarm, c(1, 1, 0, 0, 1))   # 2 sensors + master alive: alarms
phi(alarm, c(1, 1, 1, 1, 0))   # all sensors but no master: silent

# Minimal cut sets (derived from phi via the Berge transversal)
min_cuts(alarm)

# Critical states for component j
critical_states(alarm, 5L)  # the master: critical in every state where >= k sensors are alive

# Reliability at uniform component reliability p
reliability(alarm, 0.9)

# Distribution algebra: system survival via component composition
S <- algebraic.dist::surv(alarm)
S(c(1, 5, 10))

# Sampling: m independent component lifetimes, combined via system_lifetime
withr::with_seed(1, {
  x <- algebraic.dist::sampler(alarm)(1000)
  mean(x)
})
```

Notice that `min_cuts`, `critical_states`, `reliability`, `surv`, and
`sampler` all "just work" without being implemented for `alarm_dist`.
They composed from `ncomponents`, `component`, `phi`, and the
component-level distribution algebra.

## When to override a default

Defaults are designed for correctness across all topologies. They are
not always the fastest path. The two most common reasons to override:

**1. Closed-form distributional shortcuts.**

The default `surv.dist_structure` enumerates over component states via
the reliability polynomial, which is `O(2^m)`. If your subclass has an
analytical formula for `S_sys(t)`, override `surv` directly:

```r
surv.alarm_dist <- function(x, ...) {
  k <- x$k
  m_sens <- x$m_sensors
  sensor_surv <- lapply(seq_len(m_sens),
                        function(j) algebraic.dist::surv(component(x, j)))
  master_surv <- algebraic.dist::surv(component(x, x$m))
  function(t, ...) {
    vapply(t, function(ti) {
      # P(>= k sensors alive at ti) * P(master alive at ti).
      ps <- vapply(sensor_surv, function(S) S(ti), numeric(1L))
      sensor_p <- sum(vapply(seq.int(k, m_sens), function(sz) {
        sum(vapply(utils::combn(m_sens, sz, simplify = FALSE),
                   function(A) {
                     prod(ps[A]) * prod(1 - ps[setdiff(seq_len(m_sens), A)])
                   }, numeric(1L)))
      }, numeric(1L)))
      sensor_p * master_surv(ti)
    }, numeric(1L))
  }
}
```

**2. Topology-specific shortcuts.**

Closed-form `min_paths` saves the Berge transversal from rederiving them.
For our alarm system, every minimal path is a `(k-of-m-sensors)` subset
unioned with the master:

```r
min_paths.alarm_dist <- function(x) {
  k <- x$k
  m_sens <- x$m_sensors
  master_idx <- x$m
  lapply(utils::combn(m_sens, k, simplify = FALSE),
         function(P) sort(c(as.integer(P), master_idx)))
}
```

With `min_paths.alarm_dist` defined, `min_cuts(alarm)` becomes the
default Berge transversal but starts from a small explicit list rather
than re-deriving paths from `phi` (which it would do via the default
chain through `binary_grid`).

The general rule: **provide what you know in closed form; rely on
defaults for the rest**. Do not pre-emptively override a default unless
you have a reason; correctness across all dispatch chains is much easier
when only the primitives are subclass-specific.

## Inheriting from `coherent_dist`

If your subclass is a coherent system, you can inherit from
`coherent_dist` and skip step 3 entirely:

```{r}
parallel_of_master_and_sensors <- function(sensor_components, master_component) {
  components <- c(sensor_components, list(master_component))
  m <- length(components)
  obj <- coherent_dist(
    min_paths = list(seq_len(m - 1L), m),  # all sensors as one path; master as another
    components = components,
    m = m
  )
  class(obj) <- c("master_or_sensors_dist", class(obj))
  obj
}
```

`coherent_dist` already implements `ncomponents`, `component`, and
`min_paths` for you; you supply only the topology-specific data
(min_paths and components). The class chain automatically includes
`"coherent_dist"`, so all the usual dispatch (including
`dual.coherent_dist`, which produces a real dual `coherent_dist` instead
of a lazy wrapper) takes over.

## What the default `dual` does

`dual(x)` produces the dual structure `phi_dual(state) = 1 - phi(x, 1 - state)`.
When `x` inherits from `coherent_dist`, the `dual.coherent_dist` method
swaps minimal paths and cuts to produce a fresh `coherent_dist`. When
`x` is a generic `dist_structure` (your own subclass), the default
returns a lazy wrapper of class `"dual_of_system"`. The lazy wrapper
implements `ncomponents`, `component`, and `phi` by deferring to the
original; everything else (signatures, importance, surv) flows through
default methods exactly as for any other `dist_structure`.

If you want `dual()` on your subclass to return something more
specific, override:

```r
dual.alarm_dist <- function(x) {
  # ... return a transformed alarm_dist ...
}
```

## Cross-references

- The closed-form specializations in this package (`exp_series.R`,
  `wei_kofn.R`, etc.) are concrete examples of subclasses that override
  `surv`, `cdf`, `sampler`, `density`, and `hazard` for speed while
  inheriting the topology defaults.
- `serieshaz::dfr_dist_series` is an out-of-package implementor that
  adds a fully general dynamic-failure-rate component family while
  participating in the protocol.
- `kofn::kofn` is an inference layer built on top of `exp_kofn` /
  `wei_kofn`; it does not subclass `dist_structure` itself but uses
  the protocol heavily.

For more detail on any specific generic, see the `?phi`, `?min_paths`,
`?dual`, `?reliability`, and `?dist_structure` help pages.
